preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-10 Manuscript/manuscript.Rmd
addressed some ???, need to discuss others

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) [??? Were they threated separately?] produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by Peakachu at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2).

@@

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation [??? Shouldn't it be log2 + 1?] of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

## Establishing computational framework for assessing model performance

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) [??? Define N and p asap introducing. What is i? also define immediately. Adjust this whole paragraph accordingly. There's a term "hold-out" - isn't it how you should define it?] as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 3-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the test data. Performances were aggregated across each $i^{th}$ holdout chromosome using means and standard deviations.

@@

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* [??? We need to define what to italicize. Check below, make consistent throughout the manuscript] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 4). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 78 [??? 77 in the table?] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS, Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? Define all indices, what is N, p+1, i?]

@@

![**???[needs updating with new framework, takes longest, do last - ??? Is this current?] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

@@

Using our optimally built random forest model trained on ARROWHEAD/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

When built using ARROWHEAD ground truth boundaries at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 3) [??? Where is this info?]. This may be attributed to ARROWHEAD's boundaries frequently being separated by gaps, the effect especially pronounced near sparse areas of Hi-C contact matrices (Figure 5A) [??? I don't see it, data not shown?]. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] We also found that predicted TAD boundaries identified by *preciseTAD* were visually more likely to demarcate high concentrations of inter-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). [???add peakachu totals when ready]

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and Peakachu respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?]

@@

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?]

@@

**Additional File 1**. An example script for applying the ARROWHEAD tool to in situ Hi-C data (HIC001-HIC018) for resolutions 5kb, 10kb, 25kb, 50kb, 100kb to obtain chromosome-specific TAD boundaries on the GM12878 cell line.

@@

**Supplementary Table 2.** A complete list of the genomic annotations used to build the predictor space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

@@

**Supplementary Table 3.** Domain boundary and class imbalance summaries from SpectralTAD across resolution for both the
GM12878 and K562 cell lines respectively.

**Supplementary Table 4.** Loop boundary and class imbalance summaries from Peakachu 10 kb resolution for both the GM12878 and K562 cell lines respectively.

>>>>

Within this study we used both TAD boundary and loop boundary data derived from the publically available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD is part of the Juicer suite of tools for Hi-C data analysis and visualization. To obtain TAD boundaries we applied ARROWHEAD using downloaded primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (libraries HIC001-HIC018 and HIC069-HIC074 for GM12878 and K562 respectively) [??? Were they threated separately?;!!!fixed] produced with the MboI restriction enzyme (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). For chromatin loop boundaries, we used loops identified by Peakachu at 10 kb for the GM12878 and K562 cell lines [@salameh2019supervised]. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to the sparsity of the contact matrix at the 5 kb and 10 kb resolutions for the K562 cell line. Cell line-specific genomic annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] in the form of BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) (GRCh37/hg19 human genome assembly, Supplementary Table 2).

[???added spectralTAD]

@@

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A ($log_2$+1)-transformation [??? Shouldn't it be log2 + 1?; yes, fixed] of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

## Establishing optimal data level characteristics for TAD boundary region prediction

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was devised as a way for unbiased model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. Re-sampling was then performed on $A$ and an RF classifier was trained on $A$ using 3-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the test data. The process was repeated for each $i^{th}$ holdout chromosome and performances were aggregated using means and standard deviations.
[??? Define N and p asap introducing. What is i? also define immediately. Adjust this whole paragraph accordingly. There's a term "hold-out" - isn't it how you should define it?; !!!rewrote]

@@

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_;!!!talk about in meeting]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* [??? We need to define what to italicize. Check below, make consistent throughout the manuscript;!!!talk about in meeting] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and *plotHeatmap* tools offered by *deepTools* version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples;!!!fixed]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.!!!fixed, need to added spectralTAD]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 3). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 [??? 77 in the table?;!!!fixed] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 4). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? Define all indices, what is N, p+1, i?;!!!defined in methods, define here again?]

@@

![**???[needs updating with new framework, takes longest, do last - ??? Is this current?;talk about in meeting] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

@@

Using our optimally built random forest model trained on ARROWHEAD/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines;!!!working on it] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

When built using ARROWHEAD ground truth boundaries at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 2) [??? Where is this info?;!!!talk about in meeting]. This can be attributed to ARROWHEAD's boundaries frequently being separated by gaps, the effect especially pronounced near sparse areas of Hi-C contact matrices (Figure 5A) [??? I don't see it, data not shown?]. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] We also found that predicted TAD boundaries identified by *preciseTAD* were visually more likely to demarcate high concentrations of inter-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). [???add peakachu totals when ready]

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods;!!!added] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and Peakachu respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?;!!!can't be done for peakachu because we only have 1 resolution (10kb)]

@@

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?;!!!talk in meeting]

@@

**Additional File 1**. An example script for applying the ARROWHEAD tool to in situ Hi-C data (HIC001-HIC018) for resolutions 5kb, 10kb, 25kb, 50kb, 100kb to obtain chromosome-specific TAD boundaries on the GM12878 cell line. Prior to implementing SpectralTAD, the dump command offered by juicer [@durand2016juicer] was used to extract the contacts in sparse data matrix format.

@@

**Supplementary Table 2.** Domain boundary and class imbalance summaries from ARROWHEAD for the K562 cell line. [???add spectralTAD]

**Supplementary Table 3.** Loop boundary and class imbalance summaries from Peakachu 10 kb resolution for both the GM12878 and K562 cell lines respectively.

**Supplementary Table 4.** A complete list of the genomic annotations used to build the predictor space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

@@
